---
title: "Clean_Analyze_Death_Certificates"
author: "Wells"
date: "2025-11-17"
output: html_document
---

This exercise will involve cleaning, tabulating and mapping the death certificate data from Cheltenham and Jessup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# install.packages("htmlwidgets")
# install.packages("leaflet")
# install.packages("sf")
# install.packages("googlesheets4")
library(tidyverse)
library(rio)
library(janitor)
library(htmlwidgets)
library(leaflet)
library(sf)
library(googlesheets4)
library(tidycensus)
library(janitor)
```

### Import Extracted Death Certificate Data
```{r}
gs4_deauth()  
dc <- read_sheet("https://docs.google.com/spreadsheets/d/13CthhhfLHiPZqr1P67SCa9BVUyb5Rl6OBhUwvqawDEk/edit?usp=sharing", sheet = "responses_copy") |> 
    select(-Duplicates) |> 
  clean_names()


```

## Mapping place of origin

```{r}

dc |> 
  mutate(origin = birth_place_city_or_town_and_state_write_it_as_bowie_md) |> 
  count(origin) |> 
  arrange(desc(n))


```

### clean origin

```{r}
dc <- dc |> 
  mutate(origin = str_squish(birth_place_city_or_town_and_state_write_it_as_bowie_md)) |> 
  mutate(county_origin = case_when(
         str_detect(origin, "(?i)^md\\.?$") ~ "Maryland", 
         str_detect(origin, "(?i)baltimore county") ~ "Baltimore County",  # Check this FIRST
         str_detect(origin, "(?i)baltimore") ~ "Baltimore city",
         str_detect(origin, "(?i)easton") ~ "Talbot County",
         str_detect(origin, "(?i)frederick") ~ "Frederick County",
         str_detect(origin, "(?i)Prince George's County") ~ "Prince George's County",
         str_detect(origin, "(?i)Prince Georges County, Md.") ~ "Prince George's County",
          str_detect(origin, "(?i)Seat Pleasant") ~ "Prince George's County",
         str_detect(origin, "(?i)unknown") ~ "Unknown",
         str_detect(origin, "(?i)Centreville") ~ "Queen Anne's County",
         str_detect(origin, "(?i)church hill") ~ "Queen Anne's County",
         str_detect(origin, "(?i)Salisbury") ~ "Wicomico County",
         str_detect(origin, "(?i)Pocomoke") ~ "Worcester County",
         str_detect(origin, "(?i)Anne Arrundel County") ~ "Anne Arundel County",
          is.na(origin) ~ "Unknown",
               TRUE ~ origin
         )) |> 
      mutate(county_origin = str_remove_all(county_origin, ", Md.")) 
```


```{r}


dc |> 
  count(county_origin) |> 
  arrange(desc(n))


```

Map
```{r}

map <- get_acs(geography = "county", 
              variables = c(number_households = "B19001_001", median_income = "B19013_001"), 
               state = "MD", 
               year = 2020, geometry = TRUE) %>% 
  mutate(NAME = str_remove_all(NAME, ", Maryland")) |> 
  select(GEOID, NAME) |> 
  distinct()

#join geocoordinates with baltcity
dc_map <- dc |>  
 inner_join(map, by=c("county_origin"="NAME")) 


```

# First, convert to sf object

```{r}
dc_map <- dc_map |> 
  st_as_sf() |> 
  sf::st_transform(4326) |> 
  add_count(county_origin, name = "county_count")
```


```{r}

# Use colorFactor for categorical data
pal <- colorFactor(
  palette = "viridis",  # or "Set1", "Dark2", etc for categories
  domain = dc_map$county_count
)


leafMap <- leaflet(data = dc_map) %>%
  addProviderTiles(providers$Esri.WorldStreetMap) %>%
  addPolygons(
    color = ~pal(county_count),
    weight = 2.5,
    fillOpacity = 0.5,
    label = ~paste0("Count of inmates from ", county_origin, ": ", county_count)
  ) 
  
leafMap
```

#all the names, age in the tooltip. make a table in tibble
#rationalize the colors. min vs max distribution. by quartiles
#represent the people who dont have a specific county. color rest of the state gray and popup for the whole total
#annotation with state totals, md va, fl




# Clean names in Death Certificates
## Check for duplicate names
```{r}

dc |> 
  count(name) |> 
  filter(n > 1)

```

mark duplicates
```{r}

dupes <- dc |> 
  count(name) |> 
  filter(n > 1) |> 
  select(name) |> 
  mutate(name = str_squish(str_replace_all(name,'\\.',''))) |> 
  pull(name)

```


```{r}

dc <- dc |> 
  mutate(dupes = case_when(
    name %in% dupes ~ "x")
  ) |> 
  arrange(personal_id) |> 
  select(dupes, everything()) 


```


combine cause of death
I am combining the cause of death and if different columns unless there was a mistake
```{r}

dc <- dc |> 
  mutate(cause_death_update = paste(cause_of_death, if_cause_of_death_different_put_that_here, sep = "; ")) |> 
  mutate(cause_death_update = str_replace_all(cause_death_update, "NA", "")) |> 
  #removing pesky leading, trailing semicolons
  mutate(cause_death_update = str_remove_all(cause_death_update, "^;+|;+$")) |> 
  mutate(cause_death_update = str_remove_all(cause_death_update, "^[;\\s]+|[;\\s]+$")) |> 
  mutate(cause_death_update = str_replace_all (cause_death_update, "blank", "")) |> 
  separate(col = cause_death_update, into = c("death_cause1", "death_cause2", "death_cause3"), sep = ";", extra = "drop")

```

# Extract, tabulate death case
```{r}

death_categories <- dc |>
  pivot_longer(cols = c(death_cause1, death_cause2, death_cause3), 
               names_to = "cause_type", 
               values_to = "cause") |>
  count(cause) |>
  arrange(desc(n))
  

```

#Categorize

```{r}

death_categories <- death_categories |> 
  mutate(category = case_when(
         cause=="Pulmonary Tuberculosis" ~ "tuberculosis",
         cause=="Tuberculosis" ~ "tuberculosis",
         cause=="Pulmonary tuberculosis" ~ "tuberculosis", 
         cause=="Military tuberculosis" ~ "tuberculosis", 
         cause=="General Tuberculosis" ~ "tuberculosis", 
         cause=="Lobar Pneumonia" ~ "pneumonia",
         cause=="Pneumonia" ~ "pneumonia",
         cause=="Pneumonia Double so called" ~ "pneumonia",
         cause=="Pneumonia Lobar" ~ "pneumonia",
         TRUE ~ cause
         )) |> 
  group_by(category) |> 
  summarize(total=sum(n)) |> 
  arrange(desc(total))


```

# Notes
#-----------Notes -----------------


```{r}

dc <- dc |> 
  mutate(origin = str_squish(birth_place_city_or_town_and_state_write_it_as_bowie_md)) |> 
  mutate(origin_clean = case_when(
         str_detect(origin, "(?i)^md\\.?$") ~ "Maryland", 
         str_detect(origin, "(?i)baltimore county") ~ "Baltimore_County",  # Check this FIRST
         str_detect(origin, "(?i)baltimore") ~ "Baltimore_City",
         str_detect(origin, "(?i)easton") ~ "Easton",
         str_detect(origin, "(?i)frederick") ~ "Frederick",
         str_detect(origin, "(?i)Prince George's County") ~ "Prince_Georges_County",
         str_detect(origin, "(?i)Prince Georges County, Md.") ~ "Prince_Georges_County",
         str_detect(origin, "(?i)unknown") ~ "Unknown",
         is.na(origin) ~ "Unknown",
            TRUE ~ origin
  )) |> 
  mutate(origin_clean = str_remove_all(origin_clean, ", Md.")) |> 
    mutate(county_origin = case_when(
      origin_clean=="Baltimore_County" ~ "Baltimore County",
      origin_clean=="Baltimore_City" ~ "Baltimore city",
      origin_clean=="Easton" ~ "Talbot County",
      origin_clean=="Frederick" ~ "Frederick County",
      origin_clean=="Prince_Georges_County" ~ "Prince George's County",
      origin_clean=="Seat Pleasant" ~ "Prince George's County",
      origin_clean=="Centreville" ~ "Queen Anne's County",
      origin_clean=="Church Hill" ~ "Queen Anne's County",
      origin_clean=="Salisbury" ~ "Worcester County",
      str_detect(origin_clean, "(?i)Pocomoke") ~ "Worcester County",
      TRUE ~ origin_clean))
      
      
    
```

```{r}
leafMap <- leaflet(data = dc_map) %>%
  addProviderTiles(providers$Esri.WorldStreetMap) %>%
  addPolygons(
    color = ~pal(county_origin),
    weight = 2.5,
    smoothFactor = 0.2,
    fillOpacity = 0.5,
    label = ~paste0(name, " - Born in: ", county_origin),  # Show name and origin
    popup = ~paste0("<b>", name, "</b><br>",
                    "Birth place: ", county_origin, "<br>",
                    "Death date: ", death_date))  # Add popup for more info
  # ) %>%
  # addLegend(
  #   position = "bottomright",
  #   pal = pal,
  #   values = ~county_origin,
  #   title = "Origin of Cheltenham Inmates"
  # )

saveWidget(leafMap, "leafMap.html")
leafMap
```
<<<<<<< HEAD:Clean_Analyze_Death_Certificates_class_nov17.Rmd
=======


#--------------NOTES BELOW ----------------------------------------------------

```{r}

# Use colorFactor for categorical data
pal <- colorFactor(
  palette = "viridis",  # or "Set1", "Dark2", etc for categories
  domain = dc_map$county_count
)


leafMap <- leaflet(data = dc_map) %>%
  addProviderTiles(providers$Esri.WorldStreetMap) %>%
  addPolygons(
    color = ~pal(county_count),
    weight = 2.5,
    fillOpacity = 0.5,
    label = ~paste0("Count of inmates from ", county_origin, ": ", county_count)
  ) 
  
leafMap
#saveWidget(leafMap, "leafMap.html")
```




#---------------NOTES BELOW -----------------------------------

```{r}

dc <- dc |> 
  mutate(origin = str_squish(birth_place_city_or_town_and_state_write_it_as_bowie_md)) |> 
  mutate(origin_clean = case_when(
         str_detect(origin, "(?i)^md\\.?$") ~ "Maryland", 
         str_detect(origin, "(?i)baltimore county") ~ "Baltimore_County",  # Check this FIRST
         str_detect(origin, "(?i)baltimore") ~ "Baltimore_City",
         str_detect(origin, "(?i)easton") ~ "Easton",
         str_detect(origin, "(?i)frederick") ~ "Frederick",
         str_detect(origin, "(?i)Prince George's County") ~ "Prince_Georges_County",
         str_detect(origin, "(?i)Prince Georges County, Md.") ~ "Prince_Georges_County",
         str_detect(origin, "(?i)unknown") ~ "Unknown",
         is.na(origin) ~ "Unknown",
            TRUE ~ origin
  )) |> 
  mutate(origin_clean = str_remove_all(origin_clean, ", Md.")) |> 
    mutate(county_origin = case_when(
      origin_clean=="Baltimore_County" ~ "Baltimore County",
      origin_clean=="Baltimore_City" ~ "Baltimore city",
      origin_clean=="Easton" ~ "Talbot County",
      origin_clean=="Frederick" ~ "Frederick County",
      origin_clean=="Prince_Georges_County" ~ "Prince George's County",
      origin_clean=="Seat Pleasant" ~ "Prince George's County",
      origin_clean=="Centreville" ~ "Queen Anne's County",
      origin_clean=="Church Hill" ~ "Queen Anne's County",
      origin_clean=="Salisbury" ~ "Worcester County",
      str_detect(origin_clean, "(?i)Pocomoke") ~ "Worcester County",
      TRUE ~ origin_clean))
      
      
    
```



```{r}
leafMap <- leaflet(data = dc_map) %>%
  addProviderTiles(providers$Esri.WorldStreetMap) %>%
  addPolygons(
    color = ~pal(county_origin),
    weight = 2.5,
    smoothFactor = 0.2,
    fillOpacity = 0.5,
    label = ~paste0(name, " - Born in: ", county_origin),  # Show name and origin
    popup = ~paste0("<b>", name, "</b><br>",
                    "Birth place: ", county_origin, "<br>",
                    "Death date: ", death_date))  # Add popup for more info
  # ) %>%
  # addLegend(
  #   position = "bottomright",
  #   pal = pal,
  #   values = ~county_origin,
  #   title = "Origin of Cheltenham Inmates"
  # )

saveWidget(leafMap, "leafMap.html")
leafMap
```



>>>>>>> 453a7df6ef8554c44d435057fe3a65c55ae02632:Clean_Analyze_Death_Certificates.Rmd
