---
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---

::: {style="text-align: center; color: #1FB0C4; font-weight: bold;"}
[Baltimore Sun v. Afro American news coverage]{style="font-size: 40pt;"}
:::

April 2, 2026 <br>

> This code compares news coverage of Cheltenham in the Baltimore Sun vs the Afro-American We will use R and the Tidytext package to explore common words, analyze sentiment, and then the Quanteda package to examine keywords in the story context.

<br>

::: {style="text-align: center;"}
**Rob Wells, Ph.D.**\
Philip Merrill College of Journalism, University of Maryland.\
[robwells\@umd.edu](mailto:robwells@umd.edu){.email} \| <a href="https://www.linkedin.com/in/rob-wells-7929329/">LinkedIn</a>.\
:::

**Ela Jalil** **Akria Kyles**

## Load software

```{r message=FALSE, warning=FALSE}
#Install packages if you don't have them already
# install.packages("tidyverse")
# install.packages("tidytext")
# install.packages("quanteda")
# install.packages("readtext")
# install.packages("DT")
# install.packages("rio")
# install.packages("lubridate")
# install.packages("udpipe")
# install.packages("spacyr")
# install.packages("zoo")


library(tidyverse)
library(tidytext) 
library(quanteda)
library(readtext)
library(DT)
library(lubridate)
library(rio)
library(udpipe)
library(spacyr)
library(zoo)
```

```{r}
setwd('/Users/elajalil/Documents/Github/Jour389_689_Cheltenham')
```

<br>

#Import Combined Afro-Sun articles

```{r}
#I'm confused becase everything is called initially to Afro_sun but then it suddenly switches to afro_sun_articles
afro_sun <- read_csv("./input/afro_sun_articles_april_2.csv")
```

### Count the titles

```{r}
afro_sun |> 
  count(pubtitle) |> 
  arrange(desc(n))
```

```{r message=FALSE, warning=FALSE}

afro_sun <- afro_sun |>
  mutate(yearmo = as.yearmon(date_clean))

```

```{r}
afro_sun |> 
  group_by(pubtitle) |> 
  count(year, name = "total") |> 
ggplot(aes(x = factor(year), y = total, fill = pubtitle)) +
 geom_col() +
  theme(legend.position = "none", axis.text.x = element_text(angle=90, hjust = 1)) +
  #coord_flip() +  
  labs(title = "Distribution by Month",
       caption = "n=1030 articles. Graphic by Rob Wells. April 2 -2026",
       x = "Total",
       y = "")
```

```{r}
#doc <- afro_sun_articles |> 
  #group_by(pubtitle) |> 
  #count(document_clean)
#there is no document named afro_sun_articles and there is no column for document_clean. Changed to document_type
doc <- afro_sun |> 
  group_by(pubtitle) |> 
  count(document_type)

```

### Clean document type

```{r}
#Got rid of afro_sun_articles. Changed from document_clean to document_type
afro_sun <- afro_sun %>%
  mutate(document_type = case_when(
    document_type == "ar ticle"  ~ "article",
    document_type == "arti cle"  ~ "article",
    document_type == "artic le"  ~ "article",
    document_type == "articl e"  ~ "article",
    TRUE ~ document_type  # Fallback to the original value if no match
  ))

                     

```

# Wells ask - create a new column counting the words of clean_text

```{r}
#got rid of afro_sun_articles to afro_sun
letters <- afro_sun |> 
  filter(document_type =="letter_to_editor")



```

# Quick view of any entry by row

```{r}
wrapped_text <- strwrap(as.character(letters$full_text[1]), width = 60)
writeLines(wrapped_text, "temp_text.txt")
file.edit("temp_text.txt")

#Can put whatever row you want to read in []
#Talk about how we want to categorize these letters
#Add another column in categorization - start working on key words for the codes
```

# Bigrams

```{r}
#changed to afro_sun - maybe shouldn't be changing it to Bal_Sun, but there is also no data set called for Bal_Sun
bigrams <- afro_sun |> 
  select(clean_text) |> 
  mutate(text = str_squish(clean_text)) |> 
  mutate(text = tolower(clean_text)) |>  
  mutate(text = str_replace_all(text, "- ", "")) |> 
  unnest_tokens(bigram, text, token="ngrams", n=2 ) |> 
  separate(bigram, c("word1", "word2"), sep = " ") |> 
  filter(!word1 %in% stop_words$word) |>
  filter(!word2 %in% stop_words$word) |> 
  count(word1, word2, sort = TRUE) |> 
  filter(!is.na(word1))

bigrams
```

[Top 20 bigrams]{style="font-size: 18px; font-weight: bold;"}

```{r}

top_20_bigrams <- bigrams |> 
  top_n(20) |> 
  mutate(bigram = paste(word1, " ", word2)) |> 
  select(bigram, n)

top_20_bigrams
```

[Create a chart]{style="font-size: 18px; font-weight: bold;"}

```{r}
ggplot(top_20_bigrams, aes(x = reorder(bigram, n), y = n, fill=n)) +
  geom_bar(stat = "identity") +
  theme(legend.position = "none") +
  coord_flip() +  
  labs(title = "Top Two-Word phrases in bal_sun coverage",
       caption = "n=81 articles, 2024-2025. Graphic by Rob Wells. 2-22-2026",
       x = "Phrase",
       y = "Count of terms")
```

::: {style="text-align: center; color: #1FB0C4; font-weight: bold;"}
# Sentiment analysis
:::

[Basic]{style="font-size: 18px; font-weight: bold;"}

```{r}
#changed to afro_sun
# load sentiment dictionary
afinn <- get_sentiments("afinn")

#tokenize the dataframe, grouping by article number
sentiment <- afro_sun |> 
  select(clean_text) |> 
  mutate(article_nmbr = row_number()) |> 
  mutate(text = str_squish(clean_text)) |> 
  mutate(text = tolower(clean_text)) |>  
  mutate(text = str_replace_all(text, "- ", "")) |> 
  group_by(article_nmbr) |> 
  unnest_tokens(tokens, text) |> 
  filter(!tokens %in% stop_words$word) 

# Sentiment analysis by joining the tokenized words with the AFINN lexicon
sentiment_analysis <- sentiment |>
  inner_join(afinn, by = c("tokens"="word")) |>
  group_by(article_nmbr) |>  
  summarize(sentiment = sum(value), .groups = "drop")

# aggregate at article level, total sentiment score (Positive or Negative)
sentiment_analysis <- sentiment_analysis |>
   group_by(article_nmbr) |> 
  mutate(sentiment_type = ifelse(sentiment >= 0, "Positive", "Negative"))
```

[Sentiment By Month]{style="font-size: 18px; font-weight: bold;"}

```{r}

#changed to afro_sun from bal_sun and changed to clean_text
# Tokenize, keep yearmo
sentiment_tokens <- afro_sun |> 
  select(yearmo, clean_text) |> 
  mutate(article_nmbr = row_number()) |> 
  mutate(text = str_squish(clean_text)) |> 
  mutate(text = tolower(clean_text)) |>  
  mutate(text = str_replace_all(text, "- ", "")) |> 
  group_by(article_nmbr, yearmo) |> 
  unnest_tokens(tokens, text) |> 
  filter(!tokens %in% stop_words$word) 

# Join with AFINN and summarize
sentiment_month <- sentiment_tokens |>
  inner_join(afinn, by = c("tokens" = "word")) |>
  # Group by yearmo here so the metadata persists
  group_by(yearmo) |>  
  summarize(sentiment = sum(value), .groups = "drop") |>
  mutate(sentiment_type = ifelse(sentiment >= 0, "Positive", "Negative"))
```

[Sentiment by month, "per article"]{style="font-size: 18px; font-weight: bold;"}

```{r}
# Sentiment analysis aggregated by month with "per article" normalization
sentiment_per_article <- sentiment_tokens |>
  inner_join(afinn, by = c("tokens" = "word")) |>
  group_by(yearmo) |>  
  summarize(
    total_sentiment = sum(value),
    article_count = n_distinct(article_nmbr),
    # This is your "per capita" / per article metric
    sentiment_per_article = total_sentiment / article_count,
    .groups = "drop"
  ) |>
  mutate(sentiment_type = ifelse(sentiment_per_article >= 0, "Positive", "Negative"))

# Quick check of the results
print(sentiment_per_article)
```

[Chart sentiment by article]{style="font-size: 18px; font-weight: bold;"}

```{r}

ggplot(sentiment_per_article, aes(x = factor(yearmo), y = sentiment_per_article, fill = sentiment_per_article)) +
  geom_col() +
  # This creates a gradient from light red to dark red
  scale_fill_gradient(low = "darkred", high = "red") + 
  theme_minimal() +
  theme(
    legend.position = "none", 
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  labs(
    title = "Sentiment of bal_sun Crime News, 2024-2025",
    subtitle = "Average AFINN score per article (Normalized)",
    caption = "n=81 articles. Graphic by Rob Wells 2-22-2026",
    x = "Month",
    y = "Sentiment Score"
  )
```

[Examine Top Negative Phrases]{style="font-size: 18px; font-weight: bold;"}

```{r}
#afro_sun
negative_bigrams <- afro_sun |> 
  select(clean_text) |> 
  mutate(text = str_squish(clean_text)) |> 
  mutate(text = tolower(clean_text)) |>  
  mutate(text = str_replace_all(text, "- ", "")) |> 
  unnest_tokens(bigram, text, token="ngrams", n=2) |> 
  separate(bigram, c("word1", "word2"), sep = " ") |> 
  filter(!word1 %in% stop_words$word) |>
  filter(!word2 %in% stop_words$word) |> 
  count(word1, word2, sort = TRUE) |> 
  filter(!is.na(word1)) |> 
  mutate(bigram = paste(word1, word2)) |>
  
  # Join sentiment for word1
  left_join(afinn, by = c("word1" = "word")) |>
  rename(value1 = value) |>
  
  # Join sentiment for word2
  left_join(afinn, by = c("word2" = "word")) |>
  rename(value2 = value) |>
  
  # Combine sentiment scores
  mutate(
    total_sentiment = coalesce(value1, 0) + coalesce(value2, 0)
  ) |> 
  
  # Keep only negative bigrams
  filter(total_sentiment < 0) |>
  
  # Optional: sort by most negative
  arrange(total_sentiment)

negative_bigrams |> head(20)
```

::: {style="text-align: center; color: #1FB0C4; font-weight: bold;"}
# Speech Parsing
:::

It uses the [udpipe](https://cran.r-project.org/web/packages/udpipe/index.html) package in R

Tokenization, Parts of Speech Tagging, Lemmatization and Dependency Parsing with the 'UDPipe' 'NLP' Toolkit.
This natural language processing toolkit provides language-agnostic 'tokenization', 'parts of speech tagging', 'lemmatization' and 'dependency parsing' of raw text.

This code loads an English language model to permit tagging of parts of speech.
<br> This version only deals with adjectives.
<br>\
It then counts all adjectives across all articles and then counts adjectives by article.
<br> Credit to Claude.ai for providing the basis of this code.
<br>

[Load English language model]{style="font-size: 18px; font-weight: bold;"}

```{r include=FALSE}
udmodel <- udpipe_download_model(language = "english")
udmodel <- udpipe_load_model(udmodel$file_model)
```

[Annotate parts of speech on articles]{style="font-size: 18px; font-weight: bold;"}

```{r}
# Then run your annotation
annotated_text <- udpipe_annotate(udmodel, 
                                 x = bal_sun$text, 
                                 doc_id = bal_sun$index,
                                   parser = "default",
                                   tagger = "default")
annotated_df <- as.data.frame(annotated_text)
```

The annotated_df columns follow a "CoNLL-U" format for annotated linguistic data in Universal Dependencies.
It permits analysis of grammatical and syntactic structure of text.

Breaking down the details in an annotated dataframe from UDPipe:

-   doc_id: Identifier for the document - this matches harris_articles_df\$filename.

-   paragraph_id: Identifier for paragraphs within each document.

-   sentence_id: Identifier for sentences within each document.

-   sentence: The full text of the sentence.

-   token_id: Identifier for each token (word, punctuation) within a sentence.

-   token: The actual word or punctuation mark.

-   lemma: The base or dictionary form of the token (e.g., "running" → "run").\
    upos: Universal Part of Speech tag - a standardized grammatical category (e.g., NOUN, VERB, ADJ).

-   xpos: Language-specific part of speech tag - more detailed than upos, varies by language.
    feats: Morphological features - additional grammatical information like Number=Sing, Tense=Past.

-   head_token_id: The ID of the token that the current token depends on in the dependency parse.

-   dep_rel: Dependency relation - describes the syntactic relationship between a token and its head (e.g., "nsubj" for nominal subject).

-   deps: Enhanced dependency graph in the format of head-relation pairs.

-   misc: Miscellaneous information that doesn't fit elsewhere, often contains spans of tokens or other metadata.

Learn more about [Universal Dependencies format](https://universaldependencies.org/format.html)

A nerdy paper about this process [Tokenizing, POS Tagging, Lemmatizing and Parsing UD 2.0 with UDPipe](https://ufal.mff.cuni.cz/~straka/papers/2017-conll_udpipe.pdf)

[Count adjectives]{style="font-size: 18px; font-weight: bold;"}

```{r}
# Filter for adjectives (upos = "ADJ") and count frequencies
adjective_counts <- annotated_df |>
  filter(upos == "ADJ") |>
  group_by(token) |>
  summarise(
    frequency = n(),
    articles = n_distinct(doc_id)
  ) |>
  arrange(desc(frequency))

datatable(adjective_counts,
          caption = "Top adjectives in bal_sun crime coverage",
          options = list(pageLength = 10)) 

```

[Interesting results:]{style="font-size: 18px; font-weight: bold;"}

-   violent is #3, appears 101 times in 40 articles. Is this racialized?
-   criminal is #4, appears 79 times in 28 articles.
-   illegal is #16, appears 35 times in 16 articles. Is this a focus on "illegal activity" or "illegal behavior"? Or "illegal drugs"?
-   safer is #21, appears 27 times in 15 articles. Is this a focus on "safer communities" or "safer neighborhoods"? Or "safer streets"?

[Extract adjective-noun pairs]{style="font-size: 18px; font-weight: bold;"}

This section extracts adjective-noun pairs where the adjective directly modifies the noun.
It then tabulates the pairs by frequency and by article.

```{r echo=TRUE, message=FALSE, warning=FALSE}
adj_noun_pairs <- annotated_df |>
  inner_join(
    annotated_df |> select(doc_id, sentence_id, token, token_id, upos),
    by = c("doc_id", "sentence_id", "head_token_id" = "token_id")
  ) |>
  # Filter for adjectives modifying nouns
  filter(
    dep_rel == "amod" &    # amod = adjectival modifier
    upos.x == "ADJ" &      # first token is adjective
    upos.y == "NOUN"       # head word is noun
  ) |>
  # Select and rename relevant columns
  select(
    article_id = doc_id,
    sentence_id,
    adjective = token.x,
    noun = token.y
  )

# Count frequencies of adjective-noun pairs
adj_noun_counts <- adj_noun_pairs |>
  group_by(adjective, noun) |>
  summarise(
    frequency = n(),
    articles = n_distinct(article_id)
  ) |>
  arrange(desc(frequency))
```

```{r}
datatable(adj_noun_counts,
          caption = "Top adjectives-noun pairs in bal_sun coverage",
          options = list(pageLength = 10)) 

```

[Interesting results]{style="font-size: 18px; font-weight: bold;"}

-   violent crime, #1, 37 times in 22 articles
-   young people, #13, 9 times in 6 articles
-   public safety, #4, 21 times in 12 articles
-   ghost guns, #16, 8 times in 5 articles

::: {style="text-align: center; color: #1FB0C4; font-weight: bold;"}
# SpaCy parsing
:::

["Industrial-Strength Natural Language Processing"]{style="font-size: 18px; font-weight: bold;"}

> This is a python package that we will load in R.
> It's much more sophisticated than udpipe.

```{r}

# Initialize spacy (one-time setup required: spacy_install())
# Load English model. Other languages available
spacy_initialize(model = "en_core_web_sm")
```

```{r}
text <- data.frame(
  doc_id = bal_sun$index,
  text = bal_sun$text,
  stringsAsFactors = FALSE
)

# Parse with spacyr
parsed <- spacy_parse(text, entity = TRUE, 
                               lemma = TRUE,
                               pos = TRUE,
                               tag= TRUE,  
                              dependency = TRUE  )


```

[Spacy Adjective-noun pairs]{style="font-size: 18px; font-weight: bold;"}

```{r}

# Create adjective-noun pairs using spaCy output
spacy_adj_noun_pairs <- parsed |>
  inner_join(
    parsed |> select(doc_id, sentence_id, token, token_id, pos),
    by = c("doc_id", "sentence_id", "head_token_id" = "token_id"),
    suffix = c("_adj", "_noun")
  ) |>
  # Filter for adjectives modifying nouns
  filter(
    dep_rel == "amod" &           # amod = adjectival modifier
    pos_adj == "ADJ" &            # first token is adjective
    pos_noun == "NOUN"            # head word is noun
  ) |>
  # Select and rename relevant columns
  select(
    article_id = doc_id,
    sentence_id,
    adjective = token_adj,
    noun = token_noun
  )

# Count frequencies of adjective-noun pairs
spacy_adj_noun_counts <- spacy_adj_noun_pairs |>
  group_by(adjective, noun) |>
  summarise(
    frequency = n(),
    articles = n_distinct(article_id),
    .groups = "drop"
  ) |>
  arrange(desc(frequency))
```

[Create Datatable for spaCy output]{style="font-size: 18px; font-weight: bold;"}

```{r}

# Display results
datatable(spacy_adj_noun_counts,
          caption = "Spacy Top adjective-noun pairs in bal_sun crime coverage",
          options = list(pageLength = 10))
```

::: {style="text-align: center; color: #1FB0C4; font-weight: bold;"}
# Examine Keywords, Phrases in Context
:::

# Skip to this part - don't worry about sentiment analysis rn

<br>

[Process Text Corpus Object]{style="font-size: 18px; font-weight: bold;"}

```{r}
#change to afro_sun
# Create a unique document ID for each article
afro_sun <- afro_sun |>
  mutate(doc_id = paste0("article_", row_number()))

# Create a quanteda corpus using the text column
library(quanteda)
#Look into documentation for quanteda to figure out how to add metadata to corpus object for the quick results

my_corpus <- corpus(afro_sun, text_field = "clean_text")

# Add metadata (optional but useful)
docvars(my_corpus)$yearmo <- afro_sun$yearmo
```

[Search for the term 'media'] {style="font-size: 18px; font-weight: bold;"}

```{r}
# Tokenize corpus
my_tokens <- tokens(my_corpus)

# KWIC search
kwic_media <- kwic(my_tokens, pattern = "media", valuetype = "fixed") |> 
  as.data.frame() |>
  mutate(doc_id = docname)

kwic_beating_injury <- kwic(my_tokens, pattern = "beating", valuetype = "fixed") |> 
  as.data.frame() |>
  mutate(doc_id = docname)

#Maybe try to add which paper is reporting - see if there is a way to add back meta data fields. 

```

[Search Journalism Terms]{style="font-size: 18px; font-weight: bold;"}

```{r}
media_narratives <- kwic(my_tokens, 
                         pattern = c("media", "reporter", "newspaper", "journalist"), 
                         valuetype = "fixed") |> 
  as.data.frame() |>
  mutate(doc_id = docname)

quanteda_df_llm <- kwic(my_tokens, phrase(c("lynch|hung|hang(ed|ing)|executed?|execution|mob\\s+law|vigilance\\s+committee")), window = 20, valuetype = "regex") %>% as.data.frame() |> 
  distinct(docname,  .keep_all = TRUE)

#hanged became changed - get rid of the question mark, can adapt this - use AI, I want a regular expression that doesn't give me change or hungry. Rob used this code for lynching project. Do this sort of examination of the data. Move KWIC into code. 
```

[Create Datatable]{style="font-size: 18px; font-weight: bold;"}

```{r}

media_narratives_table <- media_narratives |> 
  select(doc_id, pre, keyword, post)

datatable(media_narratives_table,
          caption = "Journalism Terms in Context",
          options = list(pageLength = 20))

```

::: {style="text-align: center; color: #1FB0C4; font-weight: bold;"}
# Starting Your Own Project?

### "How to get a pile of documents processed for machine learning"

### Saturday, Room 205, 10:15 a.m.- 11:15 a.m.
:::

<br>

::: {style="text-align: center;"}
**Questions?**\
**Rob Wells, Ph.D.**\
Philip Merrill College of Journalism, University of Maryland.\
[robwells\@umd.edu](mailto:robwells@umd.edu){.email} \| <a href="https://www.linkedin.com/in/rob-wells-7929329/">LinkedIn</a>.\
:::

::: {style="text-align: center;"}
**Tiasia Saunders**\
Fellow, Howard Center for Investigative Journalism.\
Philip Merrill College of Journalism, University of Maryland.\
[tiasias3\@umd.edu](mailto:tiasias3@umd.edu){.email} \| <a href="https://www.linkedin.com/in/tiasia-saunders-1494671b5/">LinkedIn</a>.\

[--30--]{style="font-size: 18px; font-weight: bold;"}
:::
