---
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---

::: {style="text-align: center; color: #1FB0C4; font-weight: bold;"}
[Baltimore Sun v. Afro American news coverage]{style="font-size: 40pt;"}
:::

April 15, 2026 <br>

> This code compares news coverage of Cheltenham in the Baltimore Sun vs the Afro-American We will use R and the Tidytext package to explore common words, analyze sentiment, and then the Quanteda package to examine keywords in the story context.

<br>

::: {style="text-align: center;"}
**Rob Wells, Ph.D.**\
Philip Merrill College of Journalism, University of Maryland.\
[robwells\@umd.edu](mailto:robwells@umd.edu){.email} \| <a href="https://www.linkedin.com/in/rob-wells-7929329/">LinkedIn</a>.\
:::

**Ela Jalil** **Akria Kyles**

## Load software

```{r message=FALSE, warning=FALSE}
#Install packages if you don't have them already
# install.packages("tidyverse")
# install.packages("tidytext")
# install.packages("quanteda")
# install.packages("readtext")
# install.packages("DT")
# install.packages("rio")
# install.packages("lubridate")
# install.packages("udpipe")
# install.packages("spacyr")
# install.packages("zoo")


library(tidyverse)
library(tidytext) 
library(quanteda)
library(readtext)
library(DT)
library(lubridate)
library(rio)
library(udpipe)
library(spacyr)
library(zoo)
library(googlesheets4)
library(janitor)
```

```{r}
#setwd('/Users/elajalil/Documents/Github/Jour389_689_Cheltenham')
```

<br>

#Import Combined Afro-Sun articles

```{r}

afro_sun <- read_csv("./input/afro_sun_articles_april_9.csv")
```

#KWIC search

```{r}
#change to afro_sun
# Create a unique document ID for each article
afro_sun <- afro_sun |>
  mutate(doc_id = paste0("article_", row_number()))

# Create a quanteda corpus using the text column
library(quanteda)
#Look into documentation for quanteda to figure out how to add metadata to corpus object for the quick results

my_corpus <- corpus(afro_sun, text_field = "clean_text")

# Add metadata (optional but useful)
docvars(my_corpus)$yearmo <- afro_sun$yearmo
```

[Search for the term 'media'] {style="font-size: 18px; font-weight: bold;"}

```{r}
# Tokenize corpus
my_tokens <- tokens(my_corpus)

#Adjust the window = 20 to get more text on either side
# KWIC search
kwic_broom <- kwic(my_tokens, pattern = "broom", valuetype = "regex",window = 20) |> 
  as.data.frame() |>
  mutate(doc_id = docname)
```

### Examine a single article

```{r}
afro_sun$clean_text[35] |>
  strwrap(width = 60) |>
  writeLines("temp_text.txt")

file.edit("temp_text.txt")
```

```{r}
kwic_broom2 <- kwic(my_tokens, 
                     pattern = c("broom", "factory", "labor"), 
                     valuetype = "regex",
                     window = 20) |> 
  as.data.frame() |>
  mutate(doc_id = docname)
```

## Searches by Ela, Akira, Nina

```{r}
kwic_beating <- kwic(my_tokens, pattern = "beating", valuetype = "fixed") |> 
  as.data.frame() |>
  mutate(doc_id = docname)

#Maybe try to add which paper is reporting - see if there is a way to add back meta data fields

kwic_beating_injury <- kwic(my_tokens, 
                     pattern = c("\\bbeat(ing|en|s)?\\b", "\\bhit(ting)?\\b", "\\b(struck|strike|strikes|striking)\\b", 
                                 "\\bbludgeon(ed|ing|s)?\\b", "\\b(shot|shooting)\\b", "\\binjur(ed|y|ies|ing)\\b"), 
                     valuetype = "regex",
                     window = 20) |> 
  as.data.frame() |>
  mutate(doc_id = docname) |>
  filter(!grepl("^white$|hitude|hity|hither|architecture", keyword, ignore.case = TRUE))
#Get rid of 27 (about police brutality), would detain boys in the jail before heading to Cheltenham,29, 40, 62,65 and 67 may be repeats, 69, 71 marks abuse but not beating,80 mentions beating but would be better under slavery mention, 86, 90, 97 (includes names of boys), 99, 102 (escapes),104, 109, 122, 124 (boys beat up an elderly man and were sentenced to Cheltenham), 130, 134 (boy is "feeble minded"), 135, 139, 142, 144, 149, 153, 154, 168 (doesn't mention Cheltenham), 173 (boys parents had died and they were living on the streets),       
#118 - boy commits suicide while convict leased, boy's name is included
#a Baltimore night club waitress is anxious for MR. MUSE to reimburse her for the money she
#spent for lawyer's fees to keep his son out of Cheltenham. She can be reached at 1031 McCulloh Street (174), 178 (violating parole), 184, 188, 193,203, 210 (boy beat by his mother instead of going to Cheltenham), 214, 235, 243, 245 (rape charge against four boys for someone in the institution), 249, 263, 272, 298, 303, 304 (man was fired after eating in a white dining room), 315, 319, 325, 328 (pretty sure this is about boys beating up other boys), 344, 349, 351, 352, 353, 355, 358 (said a boy raped the toddler daughter of a white farmer he was working for), 370, 373, 378 (held up a "little white boy" doesn't give same consideration to Black children), 379,380,382,385, 393, 398, 407, 408, 413, 416, 417, 418, 426, 433, 435, 438, 449,455 (mentioned a special dispatch, did Cheltenham have a partnership with The Sun?),457 (a boy was arrested on the complaint of someone from the Charity Organization Society), 458,470,472,476,486, 487(boys escaped),489, 495,491, 497(easy for boys to escape, neighbors are unhappy),500,503,504,506(boys were sentenced to prison for beating up a guard),511(boys attempted to kill teacher to escape),516,529,531,533,543,545,546,552,557,562,564,565,566,571,572,574,584,585, 586,591,595,598,603(boy repeatedly ran away from Cheltenham),610,613,617,631,638,643(University of Maryland using Cheltenham's steady labor supply?),663, 668(State takeover process),669, 671,674,695,701,709(mention of boy working for a farmer),711,712,714,715,718,719,720,726,735,738,744,745,749,753,755,759,768,772,782,821,823,824,827,831,832(letter to the editor about firing superintendent),878,882,883,886, 898, 900, 921,928(says that children are like wild animals), 939,942,949(op ed that says she didn't believe the abuse reports),951,963,968,981(not paying unfit employees),988, 992,1000 (St Mary's can't afford to send more boys to Cheltenham),1007,1008,1010,1014,1020 (frames state takeover to be because inability to reform boys),1023,
#615 - first Sun article about abuse at the institution, 660 (mentions abuse but focuses more on the "violence of the boys")
#Not sure about 386 - a Cheltenham official hit a boy with a car but not sure if that was outside of the institution, 440 (a boy killed another boy in the institution). 453 (another case where a boy killed another boy), 855 (mentions cruelty at Cheltenham), 1026(discusses unfair labor usage of boys),1030.
#482 - a boy murders one of his employers because he doesn't get treated properly 
#figure out ways to tighten up the search. 483 (says that they made him do intense farm work and didn't pay him, and talks about how bad Cheltenham was and that the bigger boys made the little boys do whatever they wanted, were going to lynch him), 484,  
#Maybe it would be interesting to explore WHY a boy got sent to Cheltenham, I've seen passing a bad dollar, vagrancy etc. Also different way they describe the boys. 

```

Creating case_when function to add codes

```{r}
# Get the doc_id_nums from the kwic search
kwic_ids <- kwic_beating_injury |>
  mutate(doc_id_num = as.integer(str_extract(doc_id, "\\d+"))) |>
  dplyr::pull(doc_id_num) |>
  unique()

# Doc IDs to exclude
exclude_ids <- c(
  27, 29, 40, 62, 65, 67, 69, 71, 80, 86, 90, 97, 99, 102, 104, 109, 118,
  122, 124, 130, 134, 135, 139, 142, 144, 149, 153, 154, 168, 173, 174,
  178, 184, 188, 193, 203, 210, 214, 235, 243, 245, 249, 263, 272, 298,
  303, 304, 315, 319, 325, 328, 344, 349, 351, 352, 353, 355, 358, 370,
  373, 378, 379, 380, 382, 385, 393, 398, 407, 408, 413, 416, 417, 418,
  426, 433, 435, 438, 449, 455, 457, 458, 470, 472, 476, 482, 483, 484,
  486, 487, 489, 491, 495, 497, 500, 503, 504, 506, 511, 516, 529, 531,
  533, 543, 545, 546, 552, 557, 562, 564, 565, 566, 571, 572, 574, 584,
  585, 586, 591, 595, 598, 603, 610, 613, 617, 631, 638, 643, 663, 668,
  669, 671, 674, 695, 701, 709, 711, 712, 714, 715, 718, 719, 720, 726,
  735, 738, 744, 745, 749, 753, 755, 759, 768, 772, 782, 821, 823, 824,
  827, 831, 832, 878, 882, 883, 886, 898, 900, 921, 928, 939, 942, 949,
  951, 963, 968, 981, 988, 992, 1000, 1007, 1008, 1010, 1014, 1020, 1023
)

afro_sun <- afro_sun |>
  mutate(
    doc_id_num = as.integer(str_extract(doc_id, "\\d+")),
    categories = case_when(
      doc_id_num %in% kwic_ids & !doc_id_num %in% exclude_ids ~ "beating_injury",
      TRUE ~ NA_character_
    )
  )

# Quick check
afro_sun |> count(categories)
```

Checking how many articles are in the beating_injury codde

```{r}
afro_sun %>% 
  filter(categories == "beating_injury") %>% 
  count(pubtitle)
#The Afro had 39 articles related to abuse and The Sun had 21 articles 
#The Afro 39/319 = 12.2% of articles, TBS: 21/711 = 2.9% of articles

```

## Escape Search

```{r}
kwic_escape_recapture <- kwic(my_tokens,
                     pattern = c("\\bescape(d|s|ing)?\\b", "\\bescap(ee|ees)?\\b",
                                 "\\bflee(ing)?\\b", "\\bfled\\b", "\\babscond(ed|ing|s)?\\b",
                                 "\\bbroke\\s+out\\b", "\\brecaptur(ed|e|ing|es)?\\b",
                                 "\\bcaptur(ed|e|ing|es)?\\b", "\\bcaught\\b",
                                 "\\bapprehend(ed|ing|s)?\\b"),
                     valuetype = "regex",
                     window = 20) |>
  as.data.frame() |>
  mutate(doc_id = docname) |>
  filter(
    # Remove 'caught' when it's clearly not about recapture (e.g. "caught a cold", "caught fish")
    !grepl("^caught$", keyword, ignore.case = TRUE) |
      grepl("prisoner|convict|inmate|felon|fugitive|gaol|prison|custody|escaped|abscond", 
            paste(pre, post), ignore.case = TRUE),
    
    # Remove 'captured' when used in military/metaphorical sense without prison context
    !grepl("^captur", keyword, ignore.case = TRUE) |
      grepl("prisoner|convict|inmate|escaped|gaol|prison|custody|fugitive|abscond|recaptur",
            paste(pre, post), ignore.case = TRUE),
    
    # Remove 'fled' when clearly not a prison escape (e.g. "fled the country" in political context)
    !grepl("^fled$", keyword, ignore.case = TRUE) |
      grepl("prisoner|convict|inmate|gaol|prison|custody|escaped|abscond|Cheltenham",
            paste(pre, post), ignore.case = TRUE)
  )
#26(Saying boys aren't escaping),41,47(saying the boy escaped from homes and sent to Cheltenham),69(boy stole father's money and escaped from home),72(says the broom factory doesn't allow for escape),
```

```{r}
afro_sun$clean_text[73] |>
  strwrap(width = 60) |>
  writeLines("temp_text.txt")

file.edit("temp_text.txt")
```

[Search Journalism Terms]{style="font-size: 18px; font-weight: bold;"}

```{r}
media_narratives <- kwic(my_tokens, 
                         pattern = c("media", "reporter", "newspaper", "journalist"), 
                         valuetype = "fixed") |> 
  as.data.frame() |>
  mutate(doc_id = docname)

quanteda_df_llm <- kwic(my_tokens, phrase(c("lynch|hung|hang(ed|ing)|executed?|execution|mob\\s+law|vigilance\\s+committee")), window = 20, valuetype = "regex") %>% as.data.frame() |> 
  distinct(docname,  .keep_all = TRUE)

#hanged became changed - get rid of the question mark, can adapt this - use AI, I want a regular expression that doesn't give me change or hungry. Rob used this code for lynching project. Do this sort of examination of the data. Move KWIC into code. 
```

[Create Datatable]{style="font-size: 18px; font-weight: bold;"}

```{r}

media_narratives_table <- media_narratives |> 
  select(doc_id, pre, keyword, post)

datatable(media_narratives_table,
          caption = "Journalism Terms in Context",
          options = list(pageLength = 20))

```

### Count the titles

```{r}
afro_sun |> 
  count(pubtitle) |> 
  arrange(desc(n))
```

# One-time cleaning

```{r message=FALSE, warning=FALSE}
# 
# afro_sun <- afro_sun |>
#   mutate(yearmo = as.yearmon(date_clean))
# 
# #Got rid of afro_sun_articles. Changed from document_clean to document_type
# afro_sun <- afro_sun %>%
#   mutate(document_type = case_when(
#     document_type == "ar ticle"  ~ "article",
#     document_type == "arti cle"  ~ "article",
#     document_type == "artic le"  ~ "article",
#     document_type == "articl e"  ~ "article",
#     document_type == "Article"  ~ "article",
#     TRUE ~ document_type  # Fallback to the original value if no match
#   )) |> 
#   mutate(pubtitle = case_when(
#     pubtitle=="Afro-American (1893-)" ~ "Afro_American",
#     pubtitle=="The Sun (1837-)" ~ "The_Sun",
#       TRUE ~ pubtitle 
#   ))
# 
# 
# write_csv(afro_sun, ("./input/afro_sun_articles_april_9.csv"))                  

```

```{r}
afro_sun |> 
  group_by(pubtitle) |> 
  count(year, name = "total") |> 
ggplot(aes(x = factor(year), y = total, fill = pubtitle)) +
 geom_col() +
  theme(legend.position = "none", axis.text.x = element_text(angle=90, hjust = 1)) +
  #coord_flip() +  
  labs(title = "Sun vs Afro, Cheltenham Coverage by Year",
       caption = "n=1030 articles. Graphic by Rob Wells. April 9 -2026",
       x = "Total",
       y = "")
```

###Timeline

```{r}
gs4_deauth()
timeline <- read_sheet("https://docs.google.com/spreadsheets/d/1_MbLJ7RpyzBshRhrnyj122h-uKyhAaK5xYARAwwMoA0/edit?usp=sharing", sheet = "Cheltenham_Timeline") |> 
     clean_names() |> 
mutate(year = map_chr(year, ~ {
    if (is.null(.x) || length(.x) == 0) return(NA_character_) 
    return(as.character(.x))
  })) |>
  filter(cheltenham_only =="X") 
```

## Join timeline to news coverage by year

```{r}
#table with year and afro and sun article total by year
yearly <- afro_sun |> 
  group_by(pubtitle) |> 
  count(year, name = "total") |> 
  pivot_wider(names_from = pubtitle, values_from = total) |> 
  arrange(year) |> 
  mutate(year = as.character(year))

#joined with timeline
timeline_joined <- timeline |>
  right_join(yearly, by = "year") |> 
  select(year, event, Afro_American, The_Sun, everything()) |> 
  arrange(year)

# write_csv(timeline_joined, "./output/timeline_joined_april9.csv")


```

### Count document types by publication

```{r}

doc <- afro_sun |> 
  group_by(pubtitle) |> 
  count(document_type)

```

Analyze articles Afro-American (1893-) display_ad 4 - what is in these ads?

Editorials --The Afro carried 8 editorials but nothing in the Sun?
Are we sure of that?
How can we verify it.
--Pull up one of the Afro editorials and look at 1-2 months on either side

Obituaries --check Sun, Afro obituaries

# Wells ask - create a new column counting the words of clean_text

```{r}
#got rid of afro_sun_articles to afro_sun
letters <- afro_sun |> 
  filter(document_type =="letter_to_editor")



```

# Quick view of any entry by row

```{r}
wrapped_text <- strwrap(as.character(letters$full_text[35]), width = 60)
writeLines(wrapped_text, "temp_text.txt")
file.edit("temp_text.txt")

#Can put whatever row you want to read in []
#Talk about how we want to categorize these letters
#Add another column in categorization - start working on key words for the codes
```

# Bigrams

```{r}
#changed to afro_sun - maybe shouldn't be changing it to Bal_Sun, but there is also no data set called for Bal_Sun
bigrams <- afro_sun |> 
  select(clean_text) |> 
  mutate(text = str_squish(clean_text)) |> 
  mutate(text = tolower(clean_text)) |>  
  mutate(text = str_replace_all(text, "- ", "")) |> 
  unnest_tokens(bigram, text, token="ngrams", n=2 ) |> 
  separate(bigram, c("word1", "word2"), sep = " ") |> 
  filter(!word1 %in% stop_words$word) |>
  filter(!word2 %in% stop_words$word) |> 
  count(word1, word2, sort = TRUE) |> 
  filter(!is.na(word1))

bigrams
```

[Top 20 bigrams]{style="font-size: 18px; font-weight: bold;"}

```{r}

top_20_bigrams <- bigrams |> 
  top_n(20) |> 
  mutate(bigram = paste(word1, " ", word2)) |> 
  select(bigram, n)

top_20_bigrams
```

[Create a chart]{style="font-size: 18px; font-weight: bold;"}

```{r}
ggplot(top_20_bigrams, aes(x = reorder(bigram, n), y = n, fill=n)) +
  geom_bar(stat = "identity") +
  theme(legend.position = "none") +
  coord_flip() +  
  labs(title = "Top Two-Word phrases in bal_sun coverage",
       caption = "n=81 articles, 2024-2025. Graphic by Rob Wells. 2-22-2026",
       x = "Phrase",
       y = "Count of terms")
```

::: {style="text-align: center; color: #1FB0C4; font-weight: bold;"}
# Sentiment analysis
:::

[Basic]{style="font-size: 18px; font-weight: bold;"}

```{r}
#changed to afro_sun
# load sentiment dictionary
afinn <- get_sentiments("afinn")

#tokenize the dataframe, grouping by article number
sentiment <- afro_sun |> 
  select(clean_text) |> 
  mutate(article_nmbr = row_number()) |> 
  mutate(text = str_squish(clean_text)) |> 
  mutate(text = tolower(clean_text)) |>  
  mutate(text = str_replace_all(text, "- ", "")) |> 
  group_by(article_nmbr) |> 
  unnest_tokens(tokens, text) |> 
  filter(!tokens %in% stop_words$word) 

# Sentiment analysis by joining the tokenized words with the AFINN lexicon
sentiment_analysis <- sentiment |>
  inner_join(afinn, by = c("tokens"="word")) |>
  group_by(article_nmbr) |>  
  summarize(sentiment = sum(value), .groups = "drop")

# aggregate at article level, total sentiment score (Positive or Negative)
sentiment_analysis <- sentiment_analysis |>
   group_by(article_nmbr) |> 
  mutate(sentiment_type = ifelse(sentiment >= 0, "Positive", "Negative"))
```

[Sentiment By Month]{style="font-size: 18px; font-weight: bold;"}

```{r}

#changed to afro_sun from bal_sun and changed to clean_text
# Tokenize, keep yearmo
sentiment_tokens <- afro_sun |> 
  select(yearmo, clean_text) |> 
  mutate(article_nmbr = row_number()) |> 
  mutate(text = str_squish(clean_text)) |> 
  mutate(text = tolower(clean_text)) |>  
  mutate(text = str_replace_all(text, "- ", "")) |> 
  group_by(article_nmbr, yearmo) |> 
  unnest_tokens(tokens, text) |> 
  filter(!tokens %in% stop_words$word) 

# Join with AFINN and summarize
sentiment_month <- sentiment_tokens |>
  inner_join(afinn, by = c("tokens" = "word")) |>
  # Group by yearmo here so the metadata persists
  group_by(yearmo) |>  
  summarize(sentiment = sum(value), .groups = "drop") |>
  mutate(sentiment_type = ifelse(sentiment >= 0, "Positive", "Negative"))
```

[Sentiment by month, "per article"]{style="font-size: 18px; font-weight: bold;"}

```{r}
# Sentiment analysis aggregated by month with "per article" normalization
sentiment_per_article <- sentiment_tokens |>
  inner_join(afinn, by = c("tokens" = "word")) |>
  group_by(yearmo) |>  
  summarize(
    total_sentiment = sum(value),
    article_count = n_distinct(article_nmbr),
    # This is your "per capita" / per article metric
    sentiment_per_article = total_sentiment / article_count,
    .groups = "drop"
  ) |>
  mutate(sentiment_type = ifelse(sentiment_per_article >= 0, "Positive", "Negative"))

# Quick check of the results
print(sentiment_per_article)
```

[Chart sentiment by article]{style="font-size: 18px; font-weight: bold;"}

```{r}

ggplot(sentiment_per_article, aes(x = factor(yearmo), y = sentiment_per_article, fill = sentiment_per_article)) +
  geom_col() +
  # This creates a gradient from light red to dark red
  scale_fill_gradient(low = "darkred", high = "red") + 
  theme_minimal() +
  theme(
    legend.position = "none", 
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  labs(
    title = "Sentiment of bal_sun Crime News, 2024-2025",
    subtitle = "Average AFINN score per article (Normalized)",
    caption = "n=81 articles. Graphic by Rob Wells 2-22-2026",
    x = "Month",
    y = "Sentiment Score"
  )
```

[Examine Top Negative Phrases]{style="font-size: 18px; font-weight: bold;"}

```{r}
#afro_sun
negative_bigrams <- afro_sun |> 
  select(clean_text) |> 
  mutate(text = str_squish(clean_text)) |> 
  mutate(text = tolower(clean_text)) |>  
  mutate(text = str_replace_all(text, "- ", "")) |> 
  unnest_tokens(bigram, text, token="ngrams", n=2) |> 
  separate(bigram, c("word1", "word2"), sep = " ") |> 
  filter(!word1 %in% stop_words$word) |>
  filter(!word2 %in% stop_words$word) |> 
  count(word1, word2, sort = TRUE) |> 
  filter(!is.na(word1)) |> 
  mutate(bigram = paste(word1, word2)) |>
  
  # Join sentiment for word1
  left_join(afinn, by = c("word1" = "word")) |>
  rename(value1 = value) |>
  
  # Join sentiment for word2
  left_join(afinn, by = c("word2" = "word")) |>
  rename(value2 = value) |>
  
  # Combine sentiment scores
  mutate(
    total_sentiment = coalesce(value1, 0) + coalesce(value2, 0)
  ) |> 
  
  # Keep only negative bigrams
  filter(total_sentiment < 0) |>
  
  # Optional: sort by most negative
  arrange(total_sentiment)

negative_bigrams |> head(20)
```

::: {style="text-align: center; color: #1FB0C4; font-weight: bold;"}
# Speech Parsing
:::

It uses the [udpipe](https://cran.r-project.org/web/packages/udpipe/index.html) package in R

Tokenization, Parts of Speech Tagging, Lemmatization and Dependency Parsing with the 'UDPipe' 'NLP' Toolkit.
This natural language processing toolkit provides language-agnostic 'tokenization', 'parts of speech tagging', 'lemmatization' and 'dependency parsing' of raw text.

This code loads an English language model to permit tagging of parts of speech.
<br> This version only deals with adjectives.
<br>\
It then counts all adjectives across all articles and then counts adjectives by article.
<br> Credit to Claude.ai for providing the basis of this code.
<br>

[Load English language model]{style="font-size: 18px; font-weight: bold;"}

```{r include=FALSE}
udmodel <- udpipe_download_model(language = "english")
udmodel <- udpipe_load_model(udmodel$file_model)
```

[Annotate parts of speech on articles]{style="font-size: 18px; font-weight: bold;"}

```{r}
# Then run your annotation
annotated_text <- udpipe_annotate(udmodel, 
                                 x = bal_sun$text, 
                                 doc_id = bal_sun$index,
                                   parser = "default",
                                   tagger = "default")
annotated_df <- as.data.frame(annotated_text)
```

The annotated_df columns follow a "CoNLL-U" format for annotated linguistic data in Universal Dependencies.
It permits analysis of grammatical and syntactic structure of text.

Breaking down the details in an annotated dataframe from UDPipe:

-   doc_id: Identifier for the document - this matches harris_articles_df\$filename.

-   paragraph_id: Identifier for paragraphs within each document.

-   sentence_id: Identifier for sentences within each document.

-   sentence: The full text of the sentence.

-   token_id: Identifier for each token (word, punctuation) within a sentence.

-   token: The actual word or punctuation mark.

-   lemma: The base or dictionary form of the token (e.g., "running" → "run").\
    upos: Universal Part of Speech tag - a standardized grammatical category (e.g., NOUN, VERB, ADJ).

-   xpos: Language-specific part of speech tag - more detailed than upos, varies by language.
    feats: Morphological features - additional grammatical information like Number=Sing, Tense=Past.

-   head_token_id: The ID of the token that the current token depends on in the dependency parse.

-   dep_rel: Dependency relation - describes the syntactic relationship between a token and its head (e.g., "nsubj" for nominal subject).

-   deps: Enhanced dependency graph in the format of head-relation pairs.

-   misc: Miscellaneous information that doesn't fit elsewhere, often contains spans of tokens or other metadata.

Learn more about [Universal Dependencies format](https://universaldependencies.org/format.html)

A nerdy paper about this process [Tokenizing, POS Tagging, Lemmatizing and Parsing UD 2.0 with UDPipe](https://ufal.mff.cuni.cz/~straka/papers/2017-conll_udpipe.pdf)

[Count adjectives]{style="font-size: 18px; font-weight: bold;"}

```{r}
# Filter for adjectives (upos = "ADJ") and count frequencies
adjective_counts <- annotated_df |>
  filter(upos == "ADJ") |>
  group_by(token) |>
  summarise(
    frequency = n(),
    articles = n_distinct(doc_id)
  ) |>
  arrange(desc(frequency))

datatable(adjective_counts,
          caption = "Top adjectives in bal_sun crime coverage",
          options = list(pageLength = 10)) 

```

[Interesting results:]{style="font-size: 18px; font-weight: bold;"}

-   violent is #3, appears 101 times in 40 articles. Is this racialized?
-   criminal is #4, appears 79 times in 28 articles.
-   illegal is #16, appears 35 times in 16 articles. Is this a focus on "illegal activity" or "illegal behavior"? Or "illegal drugs"?
-   safer is #21, appears 27 times in 15 articles. Is this a focus on "safer communities" or "safer neighborhoods"? Or "safer streets"?

[Extract adjective-noun pairs]{style="font-size: 18px; font-weight: bold;"}

This section extracts adjective-noun pairs where the adjective directly modifies the noun.
It then tabulates the pairs by frequency and by article.

```{r echo=TRUE, message=FALSE, warning=FALSE}
adj_noun_pairs <- annotated_df |>
  inner_join(
    annotated_df |> select(doc_id, sentence_id, token, token_id, upos),
    by = c("doc_id", "sentence_id", "head_token_id" = "token_id")
  ) |>
  # Filter for adjectives modifying nouns
  filter(
    dep_rel == "amod" &    # amod = adjectival modifier
    upos.x == "ADJ" &      # first token is adjective
    upos.y == "NOUN"       # head word is noun
  ) |>
  # Select and rename relevant columns
  select(
    article_id = doc_id,
    sentence_id,
    adjective = token.x,
    noun = token.y
  )

# Count frequencies of adjective-noun pairs
adj_noun_counts <- adj_noun_pairs |>
  group_by(adjective, noun) |>
  summarise(
    frequency = n(),
    articles = n_distinct(article_id)
  ) |>
  arrange(desc(frequency))
```

```{r}
datatable(adj_noun_counts,
          caption = "Top adjectives-noun pairs in bal_sun coverage",
          options = list(pageLength = 10)) 

```

[Interesting results]{style="font-size: 18px; font-weight: bold;"}

-   violent crime, #1, 37 times in 22 articles
-   young people, #13, 9 times in 6 articles
-   public safety, #4, 21 times in 12 articles
-   ghost guns, #16, 8 times in 5 articles

::: {style="text-align: center; color: #1FB0C4; font-weight: bold;"}
# SpaCy parsing
:::

["Industrial-Strength Natural Language Processing"]{style="font-size: 18px; font-weight: bold;"}

> This is a python package that we will load in R.
> It's much more sophisticated than udpipe.

```{r}

# Initialize spacy (one-time setup required: spacy_install())
# Load English model. Other languages available
spacy_initialize(model = "en_core_web_sm")
```

```{r}
text <- data.frame(
  doc_id = bal_sun$index,
  text = bal_sun$text,
  stringsAsFactors = FALSE
)

# Parse with spacyr
parsed <- spacy_parse(text, entity = TRUE, 
                               lemma = TRUE,
                               pos = TRUE,
                               tag= TRUE,  
                              dependency = TRUE  )


```

[Spacy Adjective-noun pairs]{style="font-size: 18px; font-weight: bold;"}

```{r}

# Create adjective-noun pairs using spaCy output
spacy_adj_noun_pairs <- parsed |>
  inner_join(
    parsed |> select(doc_id, sentence_id, token, token_id, pos),
    by = c("doc_id", "sentence_id", "head_token_id" = "token_id"),
    suffix = c("_adj", "_noun")
  ) |>
  # Filter for adjectives modifying nouns
  filter(
    dep_rel == "amod" &           # amod = adjectival modifier
    pos_adj == "ADJ" &            # first token is adjective
    pos_noun == "NOUN"            # head word is noun
  ) |>
  # Select and rename relevant columns
  select(
    article_id = doc_id,
    sentence_id,
    adjective = token_adj,
    noun = token_noun
  )

# Count frequencies of adjective-noun pairs
spacy_adj_noun_counts <- spacy_adj_noun_pairs |>
  group_by(adjective, noun) |>
  summarise(
    frequency = n(),
    articles = n_distinct(article_id),
    .groups = "drop"
  ) |>
  arrange(desc(frequency))
```

[Create Datatable for spaCy output]{style="font-size: 18px; font-weight: bold;"}

```{r}

# Display results
datatable(spacy_adj_noun_counts,
          caption = "Spacy Top adjective-noun pairs in bal_sun crime coverage",
          options = list(pageLength = 10))
```

::: {style="text-align: center; color: #1FB0C4; font-weight: bold;"}
# Examine Keywords, Phrases in Context
:::

# Skip to this part - don't worry about sentiment analysis rn

<br>

[Process Text Corpus Object]{style="font-size: 18px; font-weight: bold;"}

::: {style="text-align: center; color: #1FB0C4; font-weight: bold;"}
# Starting Your Own Project?

### "How to get a pile of documents processed for machine learning"

### Saturday, Room 205, 10:15 a.m.- 11:15 a.m.
:::

<br>

::: {style="text-align: center;"}
**Questions?**\
**Rob Wells, Ph.D.**\
Philip Merrill College of Journalism, University of Maryland.\
[robwells\@umd.edu](mailto:robwells@umd.edu){.email} \| <a href="https://www.linkedin.com/in/rob-wells-7929329/">LinkedIn</a>.\
:::

::: {style="text-align: center;"}
**Tiasia Saunders**\
Fellow, Howard Center for Investigative Journalism.\
Philip Merrill College of Journalism, University of Maryland.\
[tiasias3\@umd.edu](mailto:tiasias3@umd.edu){.email} \| <a href="https://www.linkedin.com/in/tiasia-saunders-1494671b5/">LinkedIn</a>.\

[--30--]{style="font-size: 18px; font-weight: bold;"}
:::
