---
title: "Prelab_demo_Clean_Analyze_Death_Certificates"
author: "Wells"
date: "2025-11-17"
output: html_document
---

**YOUR NAME HERE**


This exercise will involve cleaning, tabulating and mapping the death certificate data from Cheltenham and Jessup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# install.packages("htmlwidgets")
# install.packages("leaflet")
# install.packages("sf")
# install.packages("googlesheets4")
library(tidyverse)
library(rio)
library(janitor)
library(htmlwidgets)
library(leaflet)
library(sf)
library(googlesheets4)
library(tidycensus)
library(janitor)
```

### Import Extracted Death Certificate Data
```{r}

gs4_deauth()  
dc <- read_sheet("https://docs.google.com/spreadsheets/d/13CthhhfLHiPZqr1P67SCa9BVUyb5Rl6OBhUwvqawDEk/edit?usp=sharing", sheet = "responses_copy") |> 
    select(-Duplicates) |> 
  clean_names()


```



## Mapping place of origin

```{r}

dc |> 
  mutate(origin = birth_place_city_or_town_and_state_write_it_as_bowie_md) |> 
  count(origin) |> 
  arrange(desc(n))


```

### Massive clean origin codeblock
```{r}

dc <- dc |> 
  mutate(origin = str_squish(birth_place_city_or_town_and_state_write_it_as_bowie_md)) |> 
  mutate(county_origin = case_when(
         str_detect(origin, "(?i)^md\\.?$") ~ "Maryland", 
         str_detect(origin, "(?i)baltimore county") ~ "Baltimore County",  # Check this FIRST
         str_detect(origin, "(?i)baltimore") ~ "Baltimore city",
          is.na(origin) ~ "Unknown",
               TRUE ~ origin
         )) |> 
      mutate(county_origin = str_remove_all(county_origin, ", Md.")) 
```

```{r}
dc |> 
  count(county_origin) |> 
  arrange(desc(n))


```

# Map
### Retrieve county polygons
```{r}

map <- get_acs(geography = "county", 
              variables = c(number_households = "B19001_001", median_income = "B19013_001"), 
               state = "MD", 
               year = 2020, geometry = TRUE) %>% 
  mutate(NAME = str_remove_all(NAME, ", Maryland")) |> 
  select(GEOID, NAME) |> 
  distinct()
```

### Join geocoordinates with death certificates

```{r}

dc_map <- dc |>  
 inner_join(map, by=c("county_origin"="NAME")) 

```

### Data to a coordinate reference system
transforms the coordinate reference system (CRS) of a spatial object. Need to do this when plotting data on web maps and need to use latitude/logitude
```{r}

dc_map <- dc_map |> 
  st_as_sf() |> 
  sf::st_transform(4326) |> 
  add_count(county_origin, name = "county_count")
```



```{r}

# Use colorFactor for categorical data
pal <- colorFactor(
  palette = "viridis",  # or "Set1", "Dark2", etc for categories
  domain = dc_map$county_count
)


leafMap <- leaflet(data = dc_map) %>%
  addProviderTiles(providers$Esri.WorldStreetMap) %>%
  addPolygons(
    color = ~pal(county_count),
    weight = 2.5,
    fillOpacity = 0.5,
    label = ~paste0("Count of inmates from ", county_origin, ": ", county_count)
  ) 
  
leafMap
```


**Your Turn**
Split up into groups. Each group will tackle various towns, counties.
Coordinate so you don't overlap.

Clean and categorize more counties. Examine the data and modify the Massive clean origin codeblock, add more to the county_origin

```{r}
dc |> 
  count(county_origin) |> 
  arrange(desc(n))


```

#------------------------------------------------------------------------------
#Part 2: Cleaning Death Certificate Data
#------------------------------------------------------------------------------


## Check for duplicate names
```{r}

dc |> 
  count(name) |> 
  filter(n > 1)

```

### create separate list of duplicates
```{r}

dupes <- dc |> 
  count(name) |> 
  filter(n > 1) |> 
  select(name) |> 
  mutate(name = str_squish(str_replace_all(name,'\\.',''))) |> 
  pull(name)

```

### mark duplicates
```{r}

dc <- dc |> 
  mutate(dupes = case_when(
    name %in% dupes ~ "x")
  ) |> 
  arrange(personal_id) |> 
  select(dupes, everything()) 


```


**Meet with students to discuss duplicates**


# Causes of death
I am combining the cause of death from different columns unless there was a mistake
From the death certificate, we are counting all elements of the cause of death for a "causes of death" metric: Immediate cause of death; Intermediate (or antecedent) cause; Underlying cause of death. 
```{r}

dc <- dc |> 
  mutate(cause_death_update = paste(cause_of_death, if_cause_of_death_different_put_that_here, sep = "; ")) |> 
  mutate(cause_death_update = str_replace_all(cause_death_update, "NA", "")) |> 
  #removing pesky leading, trailing semicolons
  mutate(cause_death_update = str_remove_all(cause_death_update, "^;+|;+$")) |> 
  mutate(cause_death_update = str_remove_all(cause_death_update, "^[;\\s]+|[;\\s]+$")) |> 
  mutate(cause_death_update = str_replace_all (cause_death_update, "blank", "")) |> 
  separate(col = cause_death_update, into = c("death_cause1", "death_cause2", "death_cause3"), sep = ";|,|and|or|&|/")

```

# Extract, tabulate death case
```{r}

death_categories <- dc |>
  pivot_longer(cols = c(death_cause1, death_cause2, death_cause3), 
               names_to = "cause_type", 
               values_to = "cause") |>
  count(cause) |>
  arrange(desc(n))
  

```

#Categorize
Split up into groups. Each group will tackle various diseases.
Coordinate so you don't overlap.

```{r}

death_categories <- death_categories |> 
  mutate(category = case_when(
         cause=="Pulmonary Tuberculosis" ~ "tuberculosis",
        **FILL IN DATA CLEANING HERE**
         )) |> 
  **COUNT THE CATEGORIES**


```

